---
title: "hw02_HeatherSummers"
output: github_document
---
```{r}
library(tidyverse)
library(gapminder)
```


## Smell test the data

#### Is it a data.frame, a matrix, a vector, a list?
```{r}
str(gapminder) #Provides a way to display the structure of the gapminder object. 
```
Gapminder is a data.frame or tibble since it is able to hold different types of variables including character data (i.e. country, continent, etc.) and numeric data (i.e. lifeExp, pop, etc.) while matrices can only use one type of variable (i.e. vectors are of the same mode). 

```{r}
typeof(gapminder)
```
In addition, gapminder is a type of list. 

#### What's its class?
```{r}
class(gapminder)
```
As shown above as well as in 'str(gapminder)', the classes are tibble, table, and data frame.

#### How many variables/columns?
```{r}
ncol(gapminder)
names(gapminder) #Provides the names for each of the variables (columns).
```
There are six variables/columns including country, continent, year, life expectancy, population, and GDP per capita.

#### How many rows/observations?
```{r}
nrow(gapminder)
```
There are 1704 rows in the gapminder data.

#### Can you get these facts about "extent" or "size" in more than one way? Can you imagine different functions being useful in different contexts?
There are several different ways that you can get information regarding the extent or size of the gapminder object. See below for some examples:
```{r}
summary(gapminder)
```
The function 'summary()' is a good way to get a statistical overview of the data (i.e. min, max, median, mean, etc.).

```{r}
dim(gapminder)
```
The function 'dim(gapminder)' provides the dimensions of the data frame, thus the number of rows and columns.

```{r}
length(gapminder)
```
The function 'length(gapminder)' provides the number of elements in a list and with the gapminder data frame each column is a list so therefore the output provides the number of columns.

The functions 'dim()' and 'length()' provide a quick way of getting basic information on the size and extent of the data. 

There are also occasions where it may be useful to get more specific information such as the number of observations per continent.
```{r}
gapminder %>%
  group_by(continent) %>%
  summarize (n = n())
```
You can also get the number of unique countries for each continent.
```{r}
gapminder %>%
  group_by(continent) %>%
  summarize(n = n(),
            n_countries = n_distinct(country))
```


#### What data type is each variable?
```{r}
lapply(gapminder, class) #In order to get the classes of each of the columns in the data frame.

str(gapminder) #Also provides a way to get the data type for each variable.
```
The country and continent variables are factors, year and population are integers, and life expectancy and GDP per capita are numeric. 


## Explore individual variables

The categorical variables I have chosen to explore are continents and countries and the quantitative variables I have chosen to explore are life expectancy and GDP per capita.

#### What are possible values (or range, whichever is appropriate) of each variable?
```{r}
summary(gapminder$continent) %>%
  length()
summary(gapminder$continent)
table(gapminder$continent)
```
The first function lets us know that there are five continents listed in the gapminder dataset and the last two functions both provide the number of elements for each of the continents.

For the 'country' variable there are 142 countries listed in the gapminder dataset.
```{r}
unique(gapminder$country) %>%
  length()
```

Below are the first five countries listed in alphabetical order.

```{r}
unique(select(gapminder, country)) %>%
  head()
```

Now looking at the quantitative variables, life expectantcy and GDP percapita. 

To get the range of life expectantcy and GDP per capita you can use the functions 'range()' or 'summary()'.
```{r}
range(gapminder$lifeExp)
summary(gapminder$lifeExp)
```
The life expectantcy data ranges from 23.60 to 82.60.
```{r}
range(gapminder$gdpPercap)
summary(gapminder$gdpPercap)
```
The GDP per capita data ranges from 241.2 to 113523.1.

You can also get the range of GDP per capita by continent.
```{r}
gapminder %>%
  group_by(continent) %>%
  summarize(minGDP=min(gdpPercap), maxGDP=max(gdpPercap))
```

#### What values are typical? What's the spread? What's the distribution? 

To get the distribution of the observations based on continent I used the function 'barplot()'.
```{r}
barplot(table(gapminder$continent))
```

Based on this graph there are more observations for Africa, followed by Asia, Europe, Americas, and then Oceania. Continents display a slightly skewed right distribution.

A histogram can be used to illustrate the distribution for life expectantcy and GDP per capita.
```{r}
hist(gapminder$lifeExp)
hist(gapminder$gdpPercap)
```

Life expectantcy displays a skewed left distribution while GDP per capita displays a skewed right distribution.


## Explore various plot types

The GDP per capita per year for each continent.
```{r}
gapminder %>%
  group_by(continent,year) %>%
  ggplot(aes(x = year, y = gdpPercap)) +
  geom_point(aes(colour = continent)) +
  labs(title = "GDP Per Capita vs Year", x = "Year", y = "GDP Per Capita")
```


Now to focus on the GDP per capita of five countries.
```{r}
gapminder %>%
    filter(country %in% c("Canada", "New Zealand","France", "Belgium", "Denmark")) %>% 
             ggplot(aes(x = year, y = gdpPercap, shape = country)) + 
                   geom_line() + geom_point(size=2) +
  labs(title = "GDP Per Capita vs Year", x = "Year", y = "GDP Per Capita")
```

The same graph can also be created the following way.
```{r}
fiveCountries <- c("Canada", "New Zealand","France", "Belgium", "Denmark")
x<-droplevels(subset(gapminder, country %in% fiveCountries))
ggplot(x, aes(x = year, y = gdpPercap, shape = country)) + geom_line() + geom_point(size=2) + labs(title = "GDP Per Capita vs Year", x = "Year", y = "GDP Per Capita")
```

In addition, the graph can be generated using colours to represent different countries instead of shape.
```{r}
gapminder %>%
    filter(country %in% c("Canada", "New Zealand","France", "Belgium", "Denmark")) %>% 
    ggplot(aes(x = year, y = gdpPercap, colour = country)) + 
    geom_line() + geom_point(size=2) +
    labs(title = "GDP Per Capita vs Year", x = "Year", y = "GDP Per Capita")
```


A boxplot of the GDP per capita of five countries.
```{r}
gapminder %>%
    filter(country %in% c("Canada", "New Zealand","France", "Belgium", "Denmark")) %>% 
             ggplot(aes(x = country, y = gdpPercap)) + geom_boxplot() + labs(title = "GDP Per Capita vs Country", x = "Country", y = "GDP Per Capita" )
```

You can also add the raw data points (seen in blue) onto the boxplot.
```{r}
gapminder %>%
    filter(country %in% c("Canada", "New Zealand","France", "Belgium", "Denmark")) %>% 
             ggplot(aes(x = country, y = gdpPercap)) + geom_boxplot() + 
  geom_jitter(position = position_jitter(width = 0.1, height = 0), alpha = 0.3, size = 2.5, colour = "blue") +
  labs(title = "GDP Per Capita vs Country", x = "Country", y = "GDP Per capita" )
```

You can also just have a stripplot (univariate scatterplot) in which only the raw data points are plotted.
```{r}
gapminder %>%
    filter(country %in% c("Canada", "New Zealand","France", "Belgium", "Denmark")) %>% 
             ggplot(aes(x = country, y = gdpPercap)) + 
  geom_jitter(position = position_jitter(width = 0.1, height = 0), alpha = 0.3, size = 3.5, colour = "grey10") + 
  stat_summary(fun.y = mean, colour = "deeppink", geom = "point", size = 5) + 
  labs(title = "GDP Per Capita vs Country", x = "Country", y = "GDP Per Capita" )
```

The mean GDP per capita for each country is shown by the pink point on the plot.

Now to create a density plot of the life expectantcies of the five countries. 
```{r}
gapminder %>%
    filter(country %in% c("Canada", "New Zealand","France", "Belgium", "Denmark")) %>% 
  ggplot(aes(x = lifeExp, fill = country)) +
  geom_density (alpha = 0.25) +
  labs (title = "Life Expectantcy Distribution", x = "Life Expectantcy", y = "Density")
```

To plot GDP per capita and life expectantcy for each continent in the year 1997 I used the function 'facet_wrap()'. 
```{r}
jPch <- 21
jDarkGray <- 'grey15'
jXlim <- c(150, 114500)
jYlim <- c(25, 100)

ggplot(subset(gapminder, year == 1997),
       aes(x = gdpPercap, y = lifeExp)) +
  scale_x_log10(limits = jXlim) + ylim(jYlim) +
  geom_point(aes(size = sqrt(pop)), pch = jPch, colour = jDarkGray, show.legend = FALSE, alpha = 0.5) +
  scale_size_continuous(range=c(1,40)) + facet_wrap(~ continent) + coord_fixed(ratio = 1/43) + aes(fill = country) + scale_fill_manual(values = country_colors) + theme_bw() + theme(strip.text = element_text(size = rel(1.1))) + labs(title = "Life Expectancy vs GDP Per Capita in 1997", 
      x = "GDP Per Capita", y = "Life Expectancy")
#The gapminder package comes with colour schemes for countries and continents, use the function 'country_colours'.
```

## But I want to do more!
```{r}
filter(gapminder, country == c("Rwanda", "Afghanistan"))
```

The output above is missing some of the data for both countries since the function is assuming that both countries are one and thus alternates years and only includes 12 observations instead of 24. Below is the correct way to get the data for Rwanda and Afghanistan.
```{r}
library(knitr)
gapminder %>%
    filter(country %in% c("Rwanda", "Afghanistan")) %>% 
  kable(digits = 1, align = 'c')
```



If we are interested in getting the minimum life expectantcy in Europe over time while retaining information about which country contributes the extreme value, the following output can be generated.
```{r}
gapminder %>%
  filter(continent == "Europe") %>% 
  select(year, country, lifeExp) %>% 
  arrange(year) %>% 
  group_by(year) %>% 
  top_n(1, wt = lifeExp) %>% 
  kable(digits = 2, align = 'c')
```

From this table we can see that over time the minimum life expectantcy in Europe has gone up with Norway having the lowest life expectantcy (72.67) in 1952 and Iceland having the lowest life expectantcy (81.76) in 2007.


## Report your process

While completing this assignment I encountered few to no frustrations with using R Markdown and GitHub which made the coding experience much more enjoyable than the first assignment. I encountered a few small issues every once in a while when I was generating the code for the section of the assignment entitled 'Exploring various plot types' but found Jenny's dplyr tutorial very helpful. If I still had troubles with a plot after consulting Jenny's document I usually searched the functions on the internet. I found creating the facet plot most challenging since there were several functions I had to look up in order to understand how to generate the plot. Although we haven't learned how to create stripplots in class I found creating those plots was fairly easy.  

