---
title: "hw05"
author: "Heather"
date: "October 18, 2017"
output: github_document
---

```{r results='hide', message=FALSE, warning=FALSE}
library(gapminder)
library(tidyverse)
library(forcats)
library(devtools)
library(knitr)
```

## Factor management
### Gapminder version:
#### Drop Oceania
In order to remove all the entries associated with Oceania I used the function `filter()`.
```{r}
gap.no.oceania <- gapminder %>%
  filter(continent != "Oceania")
table(gap.no.oceania$continent)
```
The `filter` function was able to remove the entries associated with the continent Oceania, however the level Oceania is still present in the table. Thus, in my next attempt I used the function `factor`.
```{r}
gap.drop.oceania <- factor(gap.no.oceania$continent)

table(gap.drop.oceania)
```

Another function that is able to remove the level Oceania is `droplevels`.
```{r}
gap.nolevel.oceania <- droplevels(gap.no.oceania)

table(gap.nolevel.oceania$continent)
```
Both the `factor` and `droplevels` functions were able to remove all entries associated with Oceania.

#### Reorder the levels of `country` ****or**** `continent`

I used the maximum GDP per capita for each country over the study period (i.e. 1952 - 2007) as my summary statistic. 
```{r}
gap.gdpPercap.max <- gapminder %>%
  group_by(country) %>%
  summarize(continent = unique(continent),
            lifeExp.max = max(lifeExp),
            gdpPercap.max = max(gdpPercap)) %>%
  arrange(desc(gdpPercap.max))

knitr::kable(gap.gdpPercap.max,
             col.names = c('Country','Continent','Life Expectancy(Years)','GDP Per Capita'), 
             digits = 3, 
             align ='c', 
             format = 'html',
             caption = "Maximum Life Expectancy and GDP Per Capita Between 1952 and 2007")

head(levels(gap.gdpPercap.max$country))
```
As you can see above the country factor levels are still in alphabetical order so now I must reorder the country factor levels according to the maximum GDP per capita summary.
```{r}
gap.gdpPercap.max %>%
  mutate(country = fct_reorder(country, gdpPercap.max)) %>% knitr::kable(align='c', format = 'html')
head(levels(gap.gdpPercap.max$country))

```


### Common part:

## File I/O

## Visulation design

## Writing figure to file